Deriving an Executable Model of a Data Stream Protocol from its Specification

### 2009-09-24 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
ModelBasedTestCaseGenerationDataStreamProtocol paper: added UML model, not yet finished explained part of the test structure

In section \ref{sec:test_cases_on_the_job} the execution of the test suites
is described, and the results are analyzed and compared, both from qualitative
and quantitative points of view.

In section \ref{sec:test_cases_and_the_scenarios} we examine the different 
scenarios on which test cases based on specifications can be applied, and 
the resulting benefits it is possible to obtain.

@@

The JML specifications used to annotate Java code are of varing complexities.
Some of the specifications are simple, focusing on constraints that should hold when 
calling a method (preconditions) and the constraints on the return value (a very basic form of 
postconditions).
In listing \ref{lst:jml_example_simple} a simple JML specification example is shown, the 
method \lil{getTemperature} is declared to be \lil{/*@ pure @*/}, which means that 
it cannot change the state of any object (a postcondition), and the lower and upper bound of 
its return value are defined (another postcondition).

@@

Some of the specifications are rather complex, usually focusing on properties regarding the 
behaviour of a whole object.
In listing \ref{lst:jml_example_invariant} a complex invariant 
(a property that has to be maintained, that is, that is assumed on entry and guaranteed on 
exit of each method of an object)
example is shown.

@@

The invariant is constraining the number of samples for each medium data rate streams: 
the total number of streams is \lil{Frame.ADC_CHANNELS}; the total number of frames is 
\lil{FRAMS}; the constant constraining the number of samples is 
\lil{FRAMES/Frame.ADC_CHANNELS+1}, it means that the samples contained in a frame are
fairly distributed on the channels. 
The valid samples are counted parsing all the frames contained in a packet, selecting only 
the matching valid samples.
A medium data rate stream sample is considered valid when \lil{isADCActive()} method 
returns \lil{true}.

@@

@@

>>>>

In section \ref{sec:test_cases_on_the_job} the execution of the test suites
is described, the data collected is presented.

In section \ref{sec:test_cases_retrospectives} the data collected is analyzed, 
both from a qualitative and a quantitative points of view.

In section \ref{sec:test_cases_and_the_scenarios} the different scenarios on 
which test cases based on specifications can be applied are analyzed, and 
the resulting benefits it is possible to obtain are detailed.

@@

The JML specifications used to annotate Java code are of varing complexities.
Some of the specifications are simple, focusing on constraints that should hold when 
calling a method (preconditions) and the constraints on the return value (a very basic form of 
postconditions).
In listing \ref{lst:jml_example_simple} a simple JML specification example is shown, the 
method \lil{getTemperature} is declared to be \lil{/*@ pure @*/}, which means that 
it cannot change the state of any object (a postcondition), the specification also constraints 
the return value with a lower and upper bound (another postcondition).

@@

@@

Some of the specifications are rather complex, usually focusing on properties regarding the 
behaviour of a whole object.
In listing \ref{lst:jml_example_invariant} a complex invariant 
(a property that has to be maintained, that is, that is assumed on entry and guaranteed on 
exit of each method of an object)
example is shown. 
The invariant is constraining the number of samples for each medium data rate streams: 
the total number of streams is \lil{Frame.ADC_CHANNELS}, the total number of frames is 
\lil{FRAMS}, the constant constraining the number of samples is 
\lil{FRAMES/Frame.ADC_CHANNELS+1}, meaning that the samples contained in a frame are
fairly distributed on the channels. 
The valid samples are counted parsing all the frames contained in a packet, selecting only 
the matching valid samples.
A medium data rate stream sample is considered valid when \lil{isADCActive()} method 
returns \lil{true}.

@@

The unit test cases that have been built to verify the protocol driver 
are of two kinds: hand made unit tests and unit tests automatically 
generated based on JML specifications.
The tests have been tought to be used together, complementarly. 
Despite of this premise, the tests effectiveness has been evaluated 
separately and together, the evaluation is carried out in section 
\ref{sec:test_cases_retrospectives}.
The test effectiveness evaluation has been based on three elements:
code coverage (quantitative), development help and usefulness 
(qualitative), effective results on piloting the real board 
(quantitative).

The dual testing approach was thought as a winning option when 
compared to hand based tests or completely automated tests, in testing 
a driver for a stream communication protocol with hard constraints 
on development time.
\ST stream communication protocol is characterized by a well definite 
packet structure, and a well definite semantic on the packet contents 
as well.
On the other hand there are not strong constraints on synchronization, 
since there exists two concurrent communication channels and the protocol 
is asynchronous by construction.
The lack of hard synchronization constraints reduce the advantage to use 
a classical state machine based specification.

\todo reference to protocol tests generated from a state machine based 
specification is needed

Hard development time constraints were also a precise requirement; the time 
constraints pushed us towards a specification \emph{just good enough} for
our purposes.
A \emph{just good enough} specification is simply a specification that is not 
specifing everything, but only the software parts that are considered most 
critical, or the ones that are really easy to specify with the chosen 
specification language.
In \STSB driver JML was used to specify very simple behaviours (method 
oriented) on the whole driver implementaion, and more complex behaviour on 
the driver core interfaces.
The not covered areas were addressed by hand made unit tests.

Hand made unit tests cover behaviour not covered by the automatically 
generated tests, that is, they cover the not specified behaviour; but this 
is not the only purpose they have.
Hand made unit tests are used to drive the development, using the well known 
Test Driven Development approach, as well as to drive the most error prone 
specification as well.
Tests are used to drive the specification, a hand made unit test that should 
fail if the specification is not working as expected.

The rationale is the following: a stream communication protocol is 
not completely focused on a synchronised interaction

\begin{itemize}
 \item 
\end{itemize}

@@

%=====================================================================
\section{Retrospective on test cases effectiveness}
\label{sec:test_cases_retrospectives}

The test effectiveness evaluation criteria is of two types.

\todo describe quantitative: coverage

\todo describe qualitative: errors found, help in design, ...
